Where Do Events Come From?

### 2018-08-16 paper.tex
(>) shrink text

\usepackage{xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}

@@

@@

\begin{comment}
We start with an introduction of the structured synchronous reactive
programming model of \CEU (Section~\ref{sec.ceu}).
Then, we propose to extend \CEU with asynchronous ISRs and discuss how it can
interoperate with synchronous code through an event-driven interface
(Section~\ref{sec.isrs}).
We also propose a simple static analysis to prevent race conditions between
synchronous and asynchronous code at compile time (Section~\ref{sec.shared}).
Finally, we propose a new language runtime that enters in sleep mode
automatically to save energy (Section~\ref{sec.energy}).
\end{comment}

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this might lead to race conditions at a fine grain, but might also
affect the ordering of events at a coarse grain: the effect of an earlier event
might be perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
%Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
%original application represented as the synchronous side and the environment as
%the asynchronous side, all inside the same \CEU application.
%
Our approach is to push all subtleties of asynchronous execution into device
drivers, which impersonate the environment and are responsible to emit input
events towards regular synchronous code.
%
Synchronous code preserves the atomicity and responsiveness properties,
communicating with the environment only through events, exactly as before.

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Synchronous application that turns the LED on whenever the button
           is pressed and off whenever it is unpressed.},
  label={lst.gpio.sync},
  escapechar=\%,
]
output high/low PIN_13;  // connected to LED
input  high/low PIN_02;  // connected to button
#include "gpio.ceu"      // GPIO driver

@@

The addition of asynchronous ISRs now poses real threats concerning race
conditions since they interrupt programs at arbitrary points and could possibly
share variables with synchronous code.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller might not cope with the USART 
speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
the rest of the program (hence the prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to race conditions (e.g., ln~9,21), and \CEU
raises a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts for a short period
of time.

@@

However, we do not expect that application programmers should be required to
resolve race conditions.
\CEU supports reactive abstractions (similar to coroutines) that help hiding
concurrency issues inside drivers.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now, the application (ln~29--35) invokes the abstraction (ln~33) passing a
local buffer (ln~32).
%
The driver (ln~5--27) now hides the low level interface of the previous example
(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
accesses to the shared buffer with an \code{atomic} block (ln~18--21).
%
The abstraction copies the driver buffer into the application buffer (ln~19)
up to the requested number of bytes (ln~22).
%
On the one hand the extra copying incurs a memory and runtime overhead, but on
the other hand it prevents race conditions with the ISR.

@@

The language runtime
%implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
The runtime now also has full control over the event loop and only executes
from interrupt requests.
This allows us to propose a runtime that enters in sleep mode automatically to
save energy, as depicted in Listing~\ref{lst.env}:
%
The runtime defines an input queue (ln~1) in which ISRs enqueue new events
(e.g., Listing~\ref{lst.usart.code}, ln~12).
%
In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
which starts the program, spawns the ISRs, and reaches the first \code{await}
statements in the multiple trails of the program.
%
Then, the runtime enters the infinite event loop (ln~4--11) that queries the input
queue (ln~6) to awake the program (ln~7).
%
ISRs execute asynchronously, emitting input events into the queue, which will
be queried in subsequent iterations of the loop.
%
Note that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

Microcontrollers typically support multiple levels of sleep mode, each
progressively saves more energy at the expense of keeping less functionality
active.
%
As an example, the least efficient mode of the
\emph{ATmega328P}~\cite{arduino.atmega328p} allows for
timer interrupts since it keeps its internal clock active, while the most
efficient mode turns off all peripherals and can only awake from external
interrupts (e.g., GPIO falling edge).

>>>>

%\usepackage{microtype}
\usepackage{xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}

@@

\setlength{\belowcaptionskip}{-1pt}

@@

We start with an introduction of the structured synchronous reactive
programming model of \CEU (Section~\ref{sec.ceu}).
Then, we propose to extend it with asynchronous ISRs and discuss how to
interoperate with synchronous code through an event-driven interface
(Section~\ref{sec.isrs}).
We also propose a simple static analysis to prevent race conditions between
synchronous and asynchronous code at compile time (Section~\ref{sec.shared}).
Finally, we propose a new language runtime that enters in sleep mode
automatically to save energy (Section~\ref{sec.energy}).

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this might lead to race conditions at a fine grain, but might also
affect the ordering of events at a coarse grain: the effect of an earlier event
might be perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between \CEU and
the environment even in the presence of asynchronous ISRs.
%Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
%original application represented as the synchronous side and the environment as
%the asynchronous side, all inside the same \CEU application.
%
Our approach is to push all subtleties of asynchronous execution into device
drivers, which impersonate the environment and are responsible to emit input
events towards regular synchronous code.
%
Synchronous code preserves the atomicity and responsiveness properties,
communicating with the environment only through events, exactly as before.

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Synchronous code that turns the LED on whenever the button is
           pressed and off whenever it is unpressed.},
  label={lst.gpio.sync},
  escapechar=\%,
]
output high/low PIN_13;  // connected to LED
input  high/low PIN_02;  // connected to button
#include "gpio.ceu"      // GPIO driver

@@

The addition of asynchronous ISRs now poses real threats concerning race
conditions since they interrupt execution at arbitrary points and could
possibly share memory with synchronous code.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller might not cope with the
USART speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~10).
%
As a possible race condition, the ISR might fire and append a new byte to the
buffer (ln~9) just before the application clears it (ln~21), in which case the
new byte would be lost forever.

\CEU treats a \code{spawn~async/isr} as a block that runs in parallel with the
rest of the program (hence the prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to race conditions (e.g., ln~9,21), and \CEU
raises a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts for a short period
of time.

@@

However, we do not expect that application programmers should be required to
resolve race conditions.
\CEU provides reactive abstractions (similar to coroutines) that help hiding
concurrency issues inside drivers.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now the application (ln~29--35) invokes the abstraction (ln~33) passing a local
buffer (ln~32).
%
The driver (ln~5--27) now hides the low-level interface (ln~7--8) and
implements the abstraction (ln~15--27) that protects the accesses to the shared
buffer with an \code{atomic} block (ln~18--21).
%
The code copies the driver buffer into the application buffer (ln~19) up to the
requested number of bytes (ln~22).
%
On the one hand the extra copying incurs a memory and runtime overhead, but on
the other hand it prevents race conditions with the ISR.

@@

The language runtime now
%implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
Since programs only execute from interrupts, the runtime can now also enter in
sleep mode automatically to save energy, as depicted in Listing~\ref{lst.env}:
%
It first defines an input queue (ln~1) in which ISRs enqueue new events (e.g.,
Listing~\ref{lst.usart.code}, ln~12).
%
The \code{main} function generates thes \emph{boot reaction} once to start the
program (ln~3), which will spawn the ISRs and reach the first \code{await}
statements in the multiple trails of the program.
%
Then, the runtime enters the event loop (ln~4--11), which queries the input
queue (ln~6) to awake the program (ln~7).
%
ISRs execute asynchronously, emitting input events into the queue, which will
be queried in subsequent iterations of the loop.
%
Note that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

Microcontrollers typically support multiple levels of sleep modes, each
progressively saves more energy at the expense of keeping less functionality
active.
%
As an example, the least efficient mode of the
\emph{ATmega328P}~\cite{arduino.atmega328p} allows for
timer interrupts since it keeps its internal clock active, while the most
efficient mode turns off all peripherals and can only awake from external
interrupts (e.g., GPIO falling edge).
